Ethereum: A Secure Decentralised Generalised Transaction Ledger \\ {\smaller \textbf{Proof-of-Concept V}}

### 2014-05-02 Paper.tex
Most of PoC-5.

@@

If the latter two fields are missing, the node represents a simple (non-contract) account: it maintains no state and executes no code on receipt of a transaction.

@@

Where $v$ is the account validity function:
\begin{eqnarray}
\quad v(x) & \equiv & x_n \in \mathbb{P} \wedge x_b \in \mathbb{P} \wedge (x_s = x_c = \varnothing \vee w(x)) \\
\quad w(x) & \equiv & x_s \in \mathbb{B}_{32} \quad \wedge \quad x_c \in \mathbb{B}_{32}
\end{eqnarray}

@@

\begin{description}
\item[body] An unlimited size byte array specifying the EVM-code for the contract, formally $T_\mathbf{b}$.
\item[init] An unlimited size byte array specifying the EVM-code for the contract initialisation procedure, formally $T_\mathbf{i}$.
\end{description}

Both \textbf{body} and \textbf{init} are EVM-code fragments; \textbf{body} is executed each time the contract account receives a message call (either through a transaction or due to the internal execution of code). \textbf{init} is a piece of code that is executed only once at contract creation and that gets discarded immediately thereafter.

@@

The function $L_T$ prepares a transaction for being transformed into an RLP-serialised byte array:
\begin{equation}
L_T(T) \equiv \begin{cases}
(T_n, T_v, T_t, T_p, T_g, T_\mathbf{b}, T_\mathbf{i}, T_w, T_r, T_s) & \text{if} \; T_t = 0\\
(T_n, T_v, T_t, T_p, T_g, T_\mathbf{d}, T_w, T_r, T_s) & \text{otherwise} 
\end{cases}
\end{equation}

Here, we assume all components are interpreted by the RLP as integer values, with the exception of the arbitrary length byte arrays $T_\mathbf{b}$, $T_\mathbf{i}$ and $T_\mathbf{d}$ and the address hash $T_\mathbf{t}$:
\begin{equation}
\begin{array}[t]{lclclc}
T_n \in \mathbb{P} & \wedge & T_v \in \mathbb{P} & \wedge & T_p \in \mathbb{P} & \wedge \\
T_g \in \mathbb{P} & \wedge & T_w \in \mathbb{P} & \wedge & T_r \in \mathbb{P} & \wedge \\
T_s \in \mathbb{P} & \wedge & T_\mathbf{d} \in \mathbb{B} & \wedge & T_\mathbf{b} \in \mathbb{B} & \wedge \\
T_\mathbf{i} \in \mathbb{B} & \wedge & T_t \in \mathbb{B}_{20} &&\\
\end{array}
\end{equation}

@@

\begin{description}
\item[parentHash] The SHA3 256-bit hash of the parent block, in its entirety; formally $H_p$.
\item[unclesHash] The SHA3 256-bit hash of the uncles list portion of this block; formally $H_u$.
\item[coinbase] The 160-bit address to which all fees collected from the successful mining of this block be transferred; formally $H_b$.
\item[stateRoot] The SHA3 256-bit hash of the root node of the state trie, after all transactions are executed and finalisations applied; formally $H_r$.
\item[transactionsHash] The SHA3 256-bit hash of the transactions list portion of this block; formally $H_t$.
\item[difficulty] A scalar value corresponding to the difficulty level of this block. This can be calculated from the previous block's difficulty level and the timestamp; formally $H_d$.
\item[timestamp] A scalar value equal to the reasonable output of Unix's time() at this block's inception; formally $H_s$.
\item[extraData] An arbitrary byte array containing data relevant to this block. With the exception of the genesis block, this must be 32 bytes or fewer; formally $H_x$.
\item[nonce] A 256-bit hash which proves that a sufficient amount of computation has been carried out on this block; formally $H_n$.
\end{description}

The other two components in the block are simply a list of uncle block headers (of the same format as above) and a list of the transactions (as defined previously). Formally, we can refer to a block $B$:
\begin{equation}
B \equiv (B_H, B_\mathbf{T}, B_\mathbf{U})
\end{equation}

We can assert its validity if and only if it satisfies several conditions: It must be internally consistent with the uncle and transaction block hashes and the given transactions $B_\mathbf{T}$, when executed in order on the base state $\boldsymbol{\sigma}$ (derived from the final state of the parent block), result in a new state of the identity $H_r$:
\begin{equation}
\begin{array}[t]{lclc}
H_u &\equiv& \mathtt{\small SHA3}(\mathtt{\small RLP}(L_H^*(B_\mathbf{U}))) & \wedge \\
H_t &\equiv& \mathtt{\small SHA3}(\mathtt{\small RLP}(L_T^*(B_\mathbf{T}))) & \wedge \\
H_r &\equiv& \mathtt{\small TRIE}(L_S(\Pi(\boldsymbol{\sigma}, B))) &
\end{array}
\end{equation}

where:
\begin{equation}
\mathtt{\small TRIE}(L_S(\boldsymbol{\sigma})) = {P(B_H)_H}_r
\end{equation}

thus $\texttt{\small TRIE}(L_S(\boldsymbol{\sigma}))$ is the root node hash of the Merkle Patricia tree structure containing the key-value pairs of the state $\boldsymbol{\sigma}$ with values encoded using RLP, and $P(B_H)$ is the parent block of $B$, defined directly.

@@

The canonical difficulty of a block of header $H$ is defined as $D(H)$:
\begin{equation}
D(H) \equiv \begin{cases}
2^{22} & \quad \text{if genesis block}\\
{P(H)_H}_d + \lfloor\frac{{P(H)_H}_d}{1024}\rfloor & \quad \text{if $H_s < {P(H)_H}_s + 42$}\\
{P(H)_H}_d - \lfloor\frac{{P(H)_H}_d}{1024}\rfloor & \quad \text{otherwise}\\
\end{cases}
\end{equation}

@@

Thus we are able to define the block header validity function $V(H)$:
\begin{eqnarray}
V(H) & \equiv & \mathtt{PoW}(H, H_n) \leqslant \frac{2^{256}}{H_d} \quad \wedge \\
& & H_d = D(H) \quad \wedge \\
& & H_s > {P(H)_H}_s
\end{eqnarray}

@@

Thus $\boldsymbol{\sigma}'$ is the post-transactional state.

We define intrinsic gas $g_0$, the amount of gas this transaction requires to be paid prior to execution, as follows:
\begin{equation}
g_0 \equiv \begin{cases}
g_d + G_{create} & \text{if} \quad T_t = 0 \\
g_d + G_{call} & \text{otherwise} \\
\end{cases}
\end{equation}

where
\begin{equation}
g_d \equiv \begin{cases} (\lVert T_\mathbf{i}\rVert + \lVert T_\mathbf{b}\rVert) G_{data} & \text{if} \quad T_t = 0 \\
 \lVert T_\mathbf{d}\rVert G_{data} & \text{otherwise}
 \end{cases}
\end{equation}

where $\lVert T_\mathbf{d}\rVert$, $\lVert T_\mathbf{i}\rVert$ and $\lVert T_\mathbf{b}\rVert$ are the sizes, in bytes, of the transaction's associated data, initialisation and body EVM-code, respectively and $G$ is defined in Appendix \ref{app:fees}. The portion of $g_0$ that will not be paid to the miner is $g_d$, and is removed from the system.

@@

the validity is determined as:
\begin{equation}
\begin{array}[t]{rcl}
S(T) & \neq & \varnothing \quad \wedge \\
\boldsymbol{\sigma}[S(T)] & \neq & \varnothing \quad \wedge \\
T_n & = & \boldsymbol{\sigma}[S(T)]_n \quad \wedge \\
g_0 & \leq & T_g \quad \wedge \\ 
v_0 & \leq & \boldsymbol{\sigma}[S(T)]_b
\end{array}
\end{equation}

The transaction must not require more gas to be utilised in this block than the global block limit of gas, specified at $10^6$.

@@

After the message call or contract creation is processed, the state is finalised by refunding $g'$, the remaining gas, to the sender at the original rate. The Ether for the gas that was actually used is given to the miner, whose address is specified as the coinbase of the present block $B$. So we define the final state $\boldsymbol{\sigma}'$ in terms of the provisional state $\boldsymbol{\sigma}_P$:
\begin{eqnarray}
\boldsymbol{\sigma}' & \equiv & \boldsymbol{\sigma}_P \quad \text{except} \\
\boldsymbol{\sigma}'[s]_b & \equiv & \boldsymbol{\sigma}_P[s]_b + g' T_p \\
\boldsymbol{\sigma}'[m]_b & \equiv & \boldsymbol{\sigma}_P[m]_b + (T_g - g' - g_d) T_p \\
m & \equiv & {B_H}_b
\end{eqnarray}

@@

There are number of intrinsic parameters used when creating a contract: sender ($s$), nonce ($n$), available gas ($g$), gas price ($p$), endowment ($v$) together with the two arbitrary length byte arrays, $\mathbf{i}$ for the initialisation EVM code and $\mathbf{b}$ for the EVM code of the body.

We define the creation function formally as the function $\Lambda$, which evaluates from these values, together with the state $\boldsymbol{\sigma}_0$ to the tuple containing the new state, additional database entries and remaining gas $(\boldsymbol{\sigma}_P, g')$, as in section \ref{ch:transactions}:
\begin{equation}
(\boldsymbol{\sigma}_P, g') \equiv \Lambda(\boldsymbol{\sigma}_0, s, n, g, p, v, \mathbf{i}, \mathbf{b})
\end{equation}

@@

The account's nonce is initially defined as zero, the balance as the value passed, the storage as empty and the code hash as the SHA3 256-bit hash of the code, thus the mutated state becomes $\boldsymbol{\sigma}^*$:
\begin{eqnarray}
\boldsymbol{\sigma}^* & \equiv & \boldsymbol{\sigma} \quad \text{except:} \\
\boldsymbol{\sigma}^*[A(s, \boldsymbol{\sigma}[s]_n)] & \equiv & ( 0, v, \varnothing, \mathtt{\tiny SHA3}(\mathbf{b}) )
\end{eqnarray}

@@

If such an exception does not occur, then the remaining gas is refunded to the originator and the now-altered state is allowed to persevere. Thus formally, we may specify the resultant state and gas as $(\boldsymbol{\sigma}_P, g')$ where:
\begin{eqnarray}
(\boldsymbol{\sigma}^{**}, g') & \equiv & \Xi(\boldsymbol{\sigma}^*, g, I) \\
\boldsymbol{\sigma}_P & \equiv & \begin{cases}
\boldsymbol{\sigma}_0 & \text{if} \quad \boldsymbol{\sigma}^{**} = \varnothing \\
\boldsymbol{\sigma}^{**} & \text{otherwise}
\end{cases} \\
I_a & \equiv & a \\
I_o & \equiv & s \\
I_p & \equiv & p \\
I_\mathbf{d} & \equiv & () \\
I_s & \equiv & s \\
I_v & \equiv & v \\
I_\mathbf{b} & \equiv & \mathbf{i}
\end{eqnarray}

$I_\mathbf{d}$ evaluates to the empty tuple. $I_H$ and $I_n$ have no special treatment and are determined from the blockchain.

@@

If the recipient account contains no code to be executed (i.e. it is not a contract), then we define the final state as being equivalent to the checkpoint state. In the case that the recipient contains code to be executed (i.e. that the account is a contract), then the contract's body code (identified as the fragment whose SHA3 hash is $\boldsymbol{\sigma}[r]_c$) is executed according to the execution model (see section \ref{ch:model}). Just as with contract creation, if the execution halts due to an exhausted gas supply, then no gas is refunded to the caller and the state is reverted to the point immediately prior to code execution.
\begin{eqnarray}
\boldsymbol{\sigma}_P & \equiv & \begin{cases}
\boldsymbol{\sigma}_1 & \text{if} \quad \boldsymbol{\sigma}_1[r]_c = \varnothing \\
\boldsymbol{\sigma}_1 & \text{if} \quad \boldsymbol{\sigma}^{**} = \varnothing \\
\boldsymbol{\sigma}^{**} & \text{otherwise}
\end{cases} \\
(\boldsymbol{\sigma}^{**}, g', \mathbf{o}) & \equiv & \Xi(\boldsymbol{\sigma}_1, g, I) \\
I_a & \equiv & a \\
I_o & \equiv & o \\
I_p & \equiv & p \\
I_\mathbf{d} & \equiv & \mathbf{d} \\
I_s & \equiv & s \\
I_v & \equiv & v \\
\text{Let} \quad \mathtt{\tiny SHA3}(I_\mathbf{b}) & = & \boldsymbol{\sigma}[r]_c
\end{eqnarray}

@@

\begin{itemize}
\item {\small SHA3} costs $G_{sha3}$ gas
\item {\small SLOAD} costs $G_{sload}$ gas
\item {\small BALANCE} costs $G_{balance}$ gas
\item {\small SSTORE} costs $d.G_{sstore}$ gas where:
\begin{itemize}
\item $d = 2$ if the new value of the storage is non-zero and the old is zero;
\item $d = 0$ if the new value of the storage is zero and the old is non-zero;
\item $d = 1$ otherwise.
\end{itemize}
\item {\small CALL} costs $G_{call}$, though additional gas may be taken for the execution of contract body code.
\item {\small CREATE} costs $G_{create}$, though additional gas may be taken for the execution of contract initialisation code.
\item {\small STOP} costs $G_{stop}$ gas
\item All other operations cost $G_{step}$ gas.
\end{itemize}

Additionally, when memory is accessed with {\small MSTORE}, {\small MSTORE8}, {\small MLOAD}, {\small RETURN}, {\small SHA3}, {\small CREATE} or {\small CALL}, the memory should be enlarged to the smallest multiple of words such that all addressed bytes now fit in it. See Appendix \ref{app:vm} for a rigorous definition of the EVM gas cost.

@@

The process of finalising a block involves three stages:

\begin{enumerate}
\item Validate (or, if mining, determine) uncles;
\item apply rewards;
\item verify (or, if mining, compute a valid) state and nonce.
\end{enumerate}

@@

@@

The application of rewards to a block involves raising the balance of the accounts of the coinbase address of the block and each uncle by a certain amount. We raise the block's coinbase account by $R_b$, the block reward, and the coinbase of each uncle by $\frac{7}{8}$ of that. Formally we define the function $\Omega$:
\begin{eqnarray}
\Omega(B, \boldsymbol{\sigma}) & \equiv & \boldsymbol{\sigma}': \boldsymbol{\sigma}' = \boldsymbol{\sigma} \quad \text{except:} \\
\boldsymbol{\sigma}'[{B_H}_b]_b & = & \boldsymbol{\sigma}[{B_H}_b]_b + R_b \\
\forall U \in B_\mathbf{U}: \quad \boldsymbol{\sigma}'[U_b]_b & = & \boldsymbol{\sigma}[U_b]_b + \frac{7}{8}R_b
\end{eqnarray}

@@

As specified at the beginning of the present work, $\Pi$ is the state-transition function, which is defined in terms of $\Omega$, the block finalisation function and $\Upsilon$, the transaction-evaluation function, both now well-defined:
\begin{eqnarray}
\Pi(\boldsymbol{\sigma}, B) & \equiv & \Omega(B, \Upsilon(\Upsilon(\boldsymbol{\sigma}, T_0), T_1) ...) \\
B & \equiv & (H, (T_0, T_1, ...), \mathbf{U})
\end{eqnarray}

@@

The genesis block is 9 items, and is specified thus:
\begin{equation}
\big( \big( 0_{256}, \mathtt{\tiny SHA3}\big(\mathtt{\tiny RLP}\big( () \big)\big), 0_{160}, 0_{256}, \mathtt{\tiny SHA3}\big(\mathtt{\tiny RLP}\big( () \big)\big), 2^{22}, 0, (), \mathtt{\tiny SHA3}\big( (42) \big) \big), (), () \big)
\end{equation}

>>>>

I define a number of useful functions throughout. One of the more common is $\ell$, which evaluates to the last item in the given sequence:

\begin{equation}
\ell(\mathbf{x}) \equiv \mathbf{x}[\lVert \mathbf{x} \rVert - 1]
\end{equation}

@@

If the latter field is the SHA3 hash of the empty string, i.e. $\boldsymbol{\sigma}[a]_c = \texttt{\small SHA3}\big(()\big)$, then the node represents a simple account, sometimes refered to as a ``non-contract'' account.

@@

Where $v$ is the account validity function:
\begin{eqnarray}
\quad v(x) & \equiv & x_n \in \mathbb{P} \wedge x_b \in \mathbb{P} \wedge x_s \in \mathbb{B}_{32} \wedge x_c \in \mathbb{B}_{32}
\end{eqnarray}

@@

\begin{description}
\item[init] An unlimited size byte array specifying the EVM-code for the contract initialisation procedure, formally $T_\mathbf{i}$.
\end{description}

\textbf{init} is an EVM-code fragment; it returns the \textbf{body}, a second fragment of code that executed each time the account receives a message call (either through a transaction or due to the internal execution of code). \textbf{init} is executed only once at contract creation and that gets discarded immediately thereafter.

@@

The function $L_T$ prepares a transaction for being transformed into an RLP-serialised byte array:
\begin{equation}
L_T(T) \equiv \begin{cases}
(T_n, T_v, T_t, T_p, T_g, T_\mathbf{i}, T_w, T_r, T_s) & \text{if} \; T_t = 0\\
(T_n, T_v, T_t, T_p, T_g, T_\mathbf{d}, T_w, T_r, T_s) & \text{otherwise} 
\end{cases}
\end{equation}

Here, we assume all components are interpreted by the RLP as integer values, with the exception of the arbitrary length byte arrays $T_\mathbf{b}$, $T_\mathbf{i}$ and $T_\mathbf{d}$ and the address hash $T_\mathbf{t}$:
\begin{equation}
\begin{array}[t]{lclclc}
T_n \in \mathbb{P} & \wedge & T_v \in \mathbb{P} & \wedge & T_p \in \mathbb{P} & \wedge \\
T_g \in \mathbb{P} & \wedge & T_w \in \mathbb{P} & \wedge & T_r \in \mathbb{P} & \wedge \\
T_s \in \mathbb{P} & \wedge & T_\mathbf{d} \in \mathbb{B} & \wedge & T_\mathbf{i} \in \mathbb{B} & \wedge \\
T_t \in \mathbb{B}_{20} &&&&\\
\end{array}
\end{equation}

@@

\begin{description}
\item[parentHash] The SHA3 256-bit hash of the parent block, in its entirety; formally $H_p$.
\item[unclesHash] The SHA3 256-bit hash of the uncles list portion of this block; formally $H_u$.
\item[coinbase] The 160-bit address to which all fees collected from the successful mining of this block be transferred; formally $H_b$.
\item[stateRoot] The SHA3 256-bit hash of the root node of the state trie, after all transactions are executed and finalisations applied; formally $H_r$.
\item[transactionsTrie] The SHA3 256-bit hash of the root node of the trie structure populated with each transaction in the transactions list portion of the block; formally $H_t$.
\item[difficulty] A scalar value corresponding to the difficulty level of this block. This can be calculated from the previous block's difficulty level and the timestamp; formally $H_d$.
\item[timestamp] A scalar value equal to the reasonable output of Unix's time() at this block's inception; formally $H_s$.
\item[number] A scalar value equal to the number of ancestor blocks. The genesis block has a number of zero; formally $H_i$.
\item[minGasPrice] A scalar value equal to the minimum price of gas a transaction must have provided in order to be sufficient for inclusion by this miner in this block; formally $H_m$.
\item[gasLimit] A scalar value equal to the current limit of gas expenditure per block; formally $H_l$.
\item[gasUsed] A scalar value equal to the total gas used in transactions in this block; formally $H_u$.
\item[extraData] An arbitrary byte array containing data relevant to this block. With the exception of the genesis block, this must be 32 bytes or fewer; formally $H_x$.
\item[nonce] A 256-bit hash which proves that a sufficient amount of computation has been carried out on this block; formally $H_n$.
\end{description}

The other two components in the block are simply a list of uncle block headers (of the same format as above) and a list of the transaction information (as defined previously). Formally, we can refer to a block $B$:
\begin{equation}
B \equiv (B_H, B_\mathbf{T}, B_\mathbf{U})
\end{equation}

We can assert its validity if and only if it satisfies several conditions: It must be internally consistent with the uncle and transaction block hashes and the given transactions $B_\mathbf{T}$, when executed in order on the base state $\boldsymbol{\sigma}$ (derived from the final state of the parent block), result in a new state of the identity $H_r$:
\begin{equation}
\begin{array}[t]{lclc}
H_u &\equiv& \mathtt{\small SHA3}(\mathtt{\small RLP}(L_H^*(B_\mathbf{U}))) & \wedge \\
H_t &\equiv& \mathtt{\small TRIE}(\{\forall i < \lVert B_\mathbf{T} \rVert: (i, B_\mathbf{T}[i])\}) & \wedge \\
H_r &\equiv& \mathtt{\small TRIE}(L_S(\Pi(\boldsymbol{\sigma}, B))) &
\end{array}
\end{equation}

where $s[i]$ equals the root of the state trie immediately after the execution of the transaction $B_\mathbf{T}[i]$, and $g[i]$ the total gas used immediately after said transaction.

% TODO: formal.

Furthermore:

\begin{equation}
\mathtt{\small TRIE}(L_S(\boldsymbol{\sigma})) = {P(B_H)_H}_r
\end{equation}

Thus $\texttt{\small TRIE}(L_S(\boldsymbol{\sigma}))$ is the root node hash of the Merkle Patricia tree structure containing the key-value pairs of the state $\boldsymbol{\sigma}$ with values encoded using RLP, and $P(B_H)$ is the parent block of $B$, defined directly.

TODO: Transaction manifest more than just transactions: $(T[i], s[i], g[i])$

TODO: Remove extrameous references to number.

@@

The canonical difficulty of a block of header $H$ is defined as $D(H)$:
\begin{equation}
D(H) \equiv \begin{cases}
2^{22} & \text{if} \quad H_i = 0\\
{P(H)_H}_d + \lfloor\frac{{P(H)_H}_d}{1024}\rfloor & \text{if} \quad H_s < {P(H)_H}_s + 42\\
{P(H)_H}_d - \lfloor\frac{{P(H)_H}_d}{1024}\rfloor & \text{otherwise}\\
\end{cases}
\end{equation}

The canonical gas limit of a block of header $H$ is defined as $L(H)$:
\begin{eqnarray}
L(H) & \equiv & \begin{cases}
10^6 & \text{if} \quad H_i = 0\\
L'(H) & \text{otherwise}
\end{cases}\\
L'(H) & \equiv & \big\lfloor \frac{1023 {P(H)_H}_l + \lfloor \frac{6}{5}{P(H)_H}_u \rfloor}{1024}\big\rfloor
\end{eqnarray}

@@

Thus we are able to define the block header validity function $V(H)$:
\begin{eqnarray}
V(H) & \equiv & \mathtt{PoW}(H, H_n) \leqslant \frac{2^{256}}{H_d} \quad \wedge \\
& & H_d = D(H) \quad \wedge \\
& & H_l = L(H) \quad \wedge \\
& & H_s > {P(H)_H}_s \quad \wedge \\
& & \lVert H_x \rVert < 1024
\end{eqnarray}

Noting additionally that \textbf{extraData} must be at most 1024 bytes.

@@

Thus $\boldsymbol{\sigma}'$ is the post-transactional state. We also define $\Upsilon^g$ to evaluate to the amount of gas used in the execution of a transaction, to be defined later.

We define intrinsic gas $g_0$, the amount of gas this transaction requires to be paid prior to execution, as follows:
\begin{equation}
g_0 \equiv \begin{cases}
\lVert T_\mathbf{i}\rVert G_{data} + G_{transaction} & \text{if} \quad T_t = 0 \\
\lVert T_\mathbf{d}\rVert G_{data} + G_{transaction} & \text{otherwise} \\
\end{cases}
\end{equation}

where $\lVert T_\mathbf{d}\rVert$ and $\lVert T_\mathbf{i}\rVert$ are the sizes, in bytes, of the transaction's associated data and initialisation EVM-code, respectively and $G$ is defined in Appendix \ref{app:fees}.

@@

The validity is determined as:
\begin{equation}
\begin{array}[t]{rcl}
S(T) & \neq & \varnothing \quad \wedge \\
\boldsymbol{\sigma}[S(T)] & \neq & \varnothing \quad \wedge \\
T_n & = & \boldsymbol{\sigma}[S(T)]_n \quad \wedge \\
g_0 & \leq & T_g \quad \wedge \\ 
v_0 & \leq & \boldsymbol{\sigma}[S(T)]_b
\end{array}
\end{equation}

The sum of the transaction gas limit and the gas utilised in this block prior must be no greater than the block's \textbf{gasLimit}, $H_l$.

% TODO: specify formally.

@@

After the message call or contract creation is processed, the state is finalised by refunding $g'$, the remaining gas, to the sender at the original rate. The Ether for the gas that was actually used is given to the miner, whose address is specified as the coinbase of the present block $B$. So we define the final state $\boldsymbol{\sigma}'$ in terms of the provisional state $\boldsymbol{\sigma}_P$:
\begin{eqnarray}
\boldsymbol{\sigma}' & \equiv & \boldsymbol{\sigma}_P \quad \text{except} \\
\boldsymbol{\sigma}'[s]_b & \equiv & \boldsymbol{\sigma}_P[s]_b + g' T_p \\
\boldsymbol{\sigma}'[m]_b & \equiv & \boldsymbol{\sigma}_P[m]_b + (T_g - g') T_p \\
m & \equiv & {B_H}_b
\end{eqnarray}

And finally specify $\Upsilon^g$, the total gas used in this transaction:
\begin{equation}
\Upsilon^g(\boldsymbol{\sigma}, T) \equiv g_0 + g'
\end{equation}

@@

There are number of intrinsic parameters used when creating a contract: sender ($s$), nonce ($n$), available gas ($g$), gas price ($p$), endowment ($v$) together with an arbitrary length byte array, $\mathbf{i}$, the initialisation EVM code.

We define the creation function formally as the function $\Lambda$, which evaluates from these values, together with the state $\boldsymbol{\sigma}_0$ to the tuple containing the new state, additional database entries and remaining gas $(\boldsymbol{\sigma}_P, g')$, as in section \ref{ch:transactions}:
\begin{equation}
(\boldsymbol{\sigma}_P, g') \equiv \Lambda(\boldsymbol{\sigma}_0, s, n, g, p, v, \mathbf{i})
\end{equation}

@@

The account's nonce is initially defined as zero, the balance as the value passed, the storage as empty and the code hash as the SHA3 256-bit hash of the empty string, thus the mutated state becomes $\boldsymbol{\sigma}^*$:
\begin{eqnarray}
\boldsymbol{\sigma}^* & \equiv & \boldsymbol{\sigma} \quad \text{except:} \\
\boldsymbol{\sigma}^*[A(s, \boldsymbol{\sigma}[s]_n)] & \equiv & \big( 0, v, \mathtt{\tiny TRIE}(\varnothing), \mathtt{\tiny SHA3}\big(()\big) \big)
\end{eqnarray}

@@

If such an exception does not occur, then the remaining gas is refunded to the originator and the now-altered state is allowed to persevere. Thus formally, we may specify the resultant state and gas as $(\boldsymbol{\sigma}_P, g')$ where:
\begin{equation}
(\boldsymbol{\sigma}^{**}, g', \mathbf{o}) \equiv \Xi(\boldsymbol{\sigma}^*, g, I) \\
\end{equation}
\begin{eqnarray}
\quad \boldsymbol{\sigma}_P & \equiv & \begin{cases}
\boldsymbol{\sigma}_0 & \text{if} \quad \mathbf{o} = () \\
\begin{array}[b]{l}
\boldsymbol{\sigma}^{**} \quad \text{except:} \\
\boldsymbol{\sigma}_P[A(s, \boldsymbol{\sigma}[s]_n)]_{\mathbf{b}} = \mathbf{o}
\end{array} & \text{otherwise}
\end{cases} \\
I_a & \equiv & a \\
I_o & \equiv & s \\
I_p & \equiv & p \\
I_\mathbf{d} & \equiv & () \\
I_s & \equiv & s \\
I_v & \equiv & v \\
I_\mathbf{b} & \equiv & \mathbf{i}
\end{eqnarray}

$I_\mathbf{d}$ evaluates to the empty tuple. $I_H$ and $I_n$ have no special treatment and are determined from the blockchain. The exception in the determination of $\boldsymbol{\sigma}_P$ dictates that the resultant byte sequence from the execution of the initialisation code specifies the final body code for the newly-created contract's account, with $\boldsymbol{\sigma}_P[A(s, \boldsymbol{\sigma}[s]_n)]_{\mathbf{b}}$ being the newly created account's body code and $\mathbf{o}$ the output byte sequence of the code execution.

\subsection{Subtleties}
Note that while the initialisation code is executing, the newly created address exists but with no intrinsic body code. Thus any message call received by it during this time causes no code to be executed. If the initialisation execution ends with a {\small SUICIDE} instruction or normal {\small STOP} code, or if the code returned is otherwise empty, then the state changes are effectively reverted, the final state being equivalent to $\boldsymbol{\sigma}_0$, the pre-execution checkpoint state.

@@

The contract's body code (identified as the fragment whose SHA3 hash is $\boldsymbol{\sigma}[r]_c$) is executed according to the execution model (see section \ref{ch:model}). Just as with contract creation, if the execution halts due to an exhausted gas supply, then no gas is refunded to the caller and the state is reverted to the point immediately prior to code execution.
\begin{eqnarray}
\boldsymbol{\sigma}_P & \equiv & \begin{cases}
\boldsymbol{\sigma}_1 & \text{if} \quad \boldsymbol{\sigma}_1[r]_c = \varnothing \\
\boldsymbol{\sigma}_1 & \text{if} \quad \boldsymbol{\sigma}^{**} = \varnothing \\
\boldsymbol{\sigma}^{**} & \text{otherwise}
\end{cases} \\
(\boldsymbol{\sigma}^{**}, g', \mathbf{o}) & \equiv & \Xi(\boldsymbol{\sigma}_1, g, I) \\
I_a & \equiv & a \\
I_o & \equiv & o \\
I_p & \equiv & p \\
I_\mathbf{d} & \equiv & \mathbf{d} \\
I_s & \equiv & s \\
I_v & \equiv & v \\
\text{Let} \quad \mathtt{\tiny SHA3}(I_\mathbf{b}) & = & \boldsymbol{\sigma}[r]_c
\end{eqnarray}

@@

\begin{itemize}
\item {\small SHA3} costs $G_{sha3}$ gas
\item {\small SLOAD} costs $G_{sload}$ gas
\item {\small BALANCE} costs $G_{balance}$ gas
\item {\small SSTORE} costs $d.G_{sstore}$ gas where:
\begin{itemize}
\item $d = 2$ if the new value of the storage is non-zero and the old is zero;
\item $d = 0$ if the new value of the storage is zero and the old is non-zero;
\item $d = 1$ otherwise.
\end{itemize}
\item {\small CALL} costs $G_{call}$, though additional gas may be taken for the execution of contract body code.
\item {\small CREATE} costs $G_{create}$, though additional gas may be taken for the execution of contract initialisation code.
\item {\small STOP} costs $G_{stop}$ gas
\item {\small SUICIDE} costs $G_{suicide}$ gas
\item All other operations cost $G_{step}$ gas.
\end{itemize}

Additionally, when memory is accessed with {\small MSTORE}, {\small MSTORE8}, {\small MLOAD}, {\small CALLDATACOPY}, {\small CODECOPY}, {\small RETURN}, {\small SHA3}, {\small CREATE} or {\small CALL}, the memory should be enlarged to the smallest multiple of words such that all addressed bytes now fit in it. See Appendix \ref{app:vm} for a rigorous definition of the EVM gas cost.

@@

The process of finalising a block involves four stages:

\begin{enumerate}
\item Validate (or, if mining, determine) uncles;
\item validate (or, if mining, determine) transactions;
\item apply rewards;
\item verify (or, if mining, compute a valid) state and nonce.
\end{enumerate}

@@

\subsection{Transaction Validation}

The given \textbf{gasUsed} and \textbf{minGasLimit} must correspond faithfully to the transactions listed. In the case of ${B_H}_m$, the minimum gas limit, all transactions included in the transaction manifest must have a gas price, ${T_0}_p$, that is at least this value:
\begin{equation}
\forall_{T \in B_\mathbf{T}} {T_0}_p \geq {B_H}_m
\end{equation}

In the case of ${B_H}_u$, the total gas used in the block, it must be equal to the accumulated gas used according to the final transaction:
\begin{equation}
{B_H}_u = \ell(\mathbf{T})_1
\end{equation}

@@

The application of rewards to a block involves raising the balance of the accounts of the coinbase address of the block and each uncle by a certain amount. We raise the block's coinbase account by $R_b$, the block reward, and the coinbase of each uncle by $\frac{7}{8}$ of that. Formally we define the function $\Omega$:
\begin{eqnarray}
\Omega(B, \boldsymbol{\sigma}) & \equiv & \boldsymbol{\sigma}': \boldsymbol{\sigma}' = \boldsymbol{\sigma} \quad \text{except:} \\
\boldsymbol{\sigma}'[{B_H}_b]_b & = & \boldsymbol{\sigma}[{B_H}_b]_b + R_b \\
\forall_{U \in B_\mathbf{U}} \quad \boldsymbol{\sigma}'[U_b]_b & = & \boldsymbol{\sigma}[U_b]_b + \frac{7}{8}R_b
\end{eqnarray}

@@

As specified at the beginning of the present work, $\Pi$ is the state-transition function, which is defined in terms of $\Omega$, the block finalisation function and $\Upsilon$, the transaction-evaluation function, both now well-defined. We begin the definition by supposing we have a series of transactions $\mathbf{T}$.

\begin{equation}
\text{Let} \quad \mathbf{T} = (\mathbf{T}[0], \mathbf{T}[1], ...)
\end{equation}

These are the transactions that comprise the block, thus:
\begin{eqnarray}
\quad \quad B & \equiv & \Big(H, \big( \mathbf{M}[0], \mathbf{M}[1], ..., \mathbf{M}[\lVert \mathbf{T}\rVert - 1]\big), \mathbf{U}\Big) \\
\quad \quad \mathbf{M}[n] & \equiv & (\mathbf{T}[n], \mathbf{s}[n], \mathbf{g}[n])
\end{eqnarray}

Here, $\mathbf{s}$ and $\mathbf{g}$ are the corresponding sequences of states and cumulative gas used for each transaction. $\mathbf{s}$ is defined simply as the state resulting from applying the corrresponding transaction to the state resulting from the previous transaction (or the block's initial state in the case of the first such transaction):
\begin{equation}
\mathbf{s}[n] = \begin{cases} \Gamma(B) & \text{if} \quad n < 0 \\ \Upsilon(\mathbf{s}[n - 1], \mathbf{T}[n]) & \text{otherwise} \end{cases}
\end{equation}

In the case of $\mathbf{g}$, we take a similar approach defining each item as the gas used in evaluatng the corresponding transaction summed with the previous item (or zero, if it's the first), giving us a running total:
\begin{equation}
\mathbf{g}[n] = \begin{cases} 0 & \text{if} \quad n < 0 \\ \Upsilon^g(\mathbf{s}[n - 1], \mathbf{T}[n]) + \mathbf{g}[n-1] & \text{otherwise} \end{cases}
\end{equation}

Finally, we define $\Pi$ as the new state given the block reward function $\Omega$ applied to the final transaction's resultant state, $\ell(B_\mathbf{T})_1$:
\begin{equation}
\Pi(\boldsymbol{\sigma}, B) \equiv \Omega(B, \ell(B_\mathbf{T})_1)
\end{equation}

@@

The genesis block is 9 items, and is specified thus:
\begin{equation}
\big( \big( 0_{256}, \mathtt{\tiny SHA3}\big(\mathtt{\tiny RLP}\big( () \big)\big), 0_{160}, 0_{256}, 0_{256}, 2^{22}, 0, (), \mathtt{\tiny SHA3}\big( (42) \big) \big), (), () \big)
\end{equation}
